---
title: "TAC/ABC Ratio Analysis in EBS and GOA"
author: "Alberto Rovellini"
date: "01/09/2025"
format: 
  html:
    self-contained: true
    code-fold: true
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This analysis explores the Total Allowable Catch (TAC) to Acceptable Biological Catch (ABC) ratio in the Eastern Bering Sea (EBS) and Gulf of Alaska (GOA) fisheries. The analysis is based on harvest specification data from AKFIN.

## Setup

Load libraries.

```{r setup}
library(tidyverse)
library(readxl)
```

## Gulf of Alaska Analysis

The data is stored in an Excel file with harvest specifications from 1986 to 2024.

```{r goa-data}
# read in GOA
goa_specs <- read_xlsx("data/GOA_harvest specs_1986-2024.xlsx", 
                       sheet = 1, 
                       range = "A3:DO131",
                       col_names = F) 

# set names for columns
yrs <- 2024:1986
tags <- paste(c("OFL","ABC","TAC"),rep(yrs,each=3),sep="_")
colnames(goa_specs) <- c("Stock","Area",tags)

# Clean up data by removing asterisks and commas
goa_specs <- goa_specs %>%
  mutate(across(!c(Stock,Area), ~{
    x <- ifelse(tolower(.) %in% c("n/a", "n/a"), NA, .)
    x <- gsub("\\*", "", x)
    x <- gsub(",", "", x)
    as.numeric(x)
  }))

# Fill stock names and reshape data
goa_specs <- goa_specs %>%
  fill(Stock, .direction = "down")

goa_specs_long <- goa_specs %>%
  pivot_longer(-c(Stock,Area), names_to = "Var_Year", values_to = "mt") %>%
  separate(col = Var_Year,
           into = c("Var", "Year"),
           sep = "_") %>%
  mutate(Year = as.numeric(Year))

# Filter for total GOA and calculate ratios
goa_specs_tot <- goa_specs_long %>% 
  filter(Area == "Total", Var %in% c("ABC","TAC")) %>%
  drop_na() %>%
  mutate(Area = "GOA")

goa_ratio <- goa_specs_tot %>%
  pivot_wider(id_cols = c(Stock,Area,Year), names_from = Var, values_from = mt) %>%
  mutate(ratio = TAC/ABC)
```

Let's visualize the TAC/ABC ratios for GOA:

```{r goa-plot}
#| fig-width: 8
#| fig-height: 8
goa_ratio %>%
  ggplot(aes(x = Year, y = ratio)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  facet_wrap(~Stock, ncol = 3) +
  labs(title = "TAC/ABC Ratios in Gulf of Alaska",
       y = "TAC/ABC Ratio",
       x = "Year")
```

## Bering Sea Analysis

Some stocks have specifications for BSAI (Bering Sea and Aleutian Islands combined), while others have separate specifications for BS and AI.

```{r bsai-data}
bsai_specs <- read_xlsx("data/BSAI_harvest specs_1986-2024.xlsx", 
                       sheet = 1, 
                       range = "A3:GO57",
                       col_names = F) 

# Set column names
yrs <- 2024:1986
tags <- paste(c("OFL","ABC","TAC", "iTAC", "CDQ"),rep(yrs,each=5),sep="_")
colnames(bsai_specs) <- c("Stock","Area",tags)

# Clean data
bsai_specs <- bsai_specs %>%
  mutate(across(!c(Stock,Area), ~{
    x <- ifelse(tolower(.) %in% c("n/a", "n/a"), NA, .)
    x <- gsub("\\*", "", x)
    x <- gsub(",", "", x)
    as.numeric(x)
  })) %>%
  fill(Stock, .direction = "down")

# Separate and rejoin data for different areas
bsai_1 <- bsai_specs %>%
  filter(Stock %in% c("Pollock","Pacific cod","Sablefish")) %>%
  filter(Area %in% c("BS","AI")) 

bsai_2 <- bsai_specs %>%
  filter(!Stock %in% c("Pollock","Pacific cod","Sablefish")) %>%
  filter(Area == "BSAI")

bsai_specs <- rbind(bsai_1,bsai_2)

# Reshape and calculate ratios
bsai_specs_long <- bsai_specs %>%
  pivot_longer(-c(Stock,Area), names_to = "Var_Year", values_to = "mt") %>%
  separate(col = Var_Year,
           into = c("Var", "Year"),
           sep = "_") %>%
  mutate(Year = as.numeric(Year))

bsai_specs_tot <- bsai_specs_long %>% 
  filter(Var %in% c("ABC","TAC")) %>%
  drop_na() 

bsai_ratio <- bsai_specs_tot %>%
  pivot_wider(id_cols = c(Stock,Area,Year), names_from = Var, values_from = mt) %>%
  mutate(ratio = TAC/ABC)
```

Visualize the BSAI ratios:

```{r bsai-plot}
#| fig-width: 8
#| fig-height: 12
bsai_ratio %>%
  ggplot(aes(x = Year, y = ratio, color = Area)) +
  geom_line(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("#2E86AB", "#F24236", "#F5B700")) +
  theme_bw() +
  facet_wrap(~Stock, ncol = 3) +
  labs(title = "TAC/ABC Ratios in Bering Sea/Aleutian Islands",
       y = "TAC/ABC Ratio",
       x = "Year")
```

## Comparing GOA and BSAI

Now let's compare the two regions for key stocks that appear in both regions (though the names of some stock complexes are not consistent between FMPs):

```{r comparison}
#| fig-width: 8
#| fig-height: 8
# Define common species
sp <- c(
  intersect(unique(goa_ratio$Stock), unique(bsai_ratio$Stock)),
  "Rock Sole",
  "Yellowfin Sole",
  "Shallow-water Flatfish",
  "Deep-water Flatfish",
  "Rex Sole"
)

# Combine and filter data
ak_ratio <- rbind(goa_ratio,bsai_ratio) %>%
  filter(Stock %in% sp)

# Create comparison plot
ak_ratio %>%
  filter(Area != "AI") %>%
  ggplot(aes(x = Year, y = ratio, color = Area)) +
  geom_line(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("#2E86AB", "#F24236", "#F5B700", "#8F7CAF")) +
  scale_y_continuous(limits = c(0,1)) +
  geom_hline(yintercept = 0.5, color = "black", linetype = "dashed") +
  theme_bw() +
  labs(x = "Year", 
       y = "TAC/ABC",
       title = "TAC/ABC Ratios Comparison Between Regions") +
  facet_wrap(~Stock, ncol = 3)
```

### Recent Trends (Post-2010)

Zooming in on the more recent period:

```{r recent-trends}
#| fig-width: 8
#| fig-height: 8
ak_ratio %>%
  filter(Year > 2010) %>%
  filter(Area != "AI") %>%
  ggplot(aes(x = Year, y = ratio, color = Area)) +
  geom_line(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("#2E86AB", "#F24236", "#F5B700", "#8F7CAF")) +
  scale_y_continuous(limits = c(0,1)) +
  geom_hline(yintercept = 0.5, color = "black", linetype = "dashed") +
  theme_bw() +
  labs(x = "Year", 
       y = "TAC/ABC",
       title = "TAC/ABC Ratios Comparison Between Regions") +
  facet_wrap(~Stock, ncol = 3)
```

## Plot TAC and ABC for the GOA

Other way to view this. For the BSAI first.

```{r raw-specs-ebs}
#| fig-width: 8
#| fig-height: 12
# for_plot_bsai <- c("Pollock", "Pacific cod" , "Sablefish", "Yellowfin Sole", "Greenland Turbot", "Arrowtooth Flounder", "Kamchatka Flounder", "Rock Sole", "Flathead Sole", "Pacific Ocean Perch", "Northern Rockfish", "Alaska Plaice")

bsai_specs_tot %>%   
  #filter(Stock %in% for_plot_bsai) %>%
  filter(Area %in% c("BS","BSAI")) %>%
  ggplot(aes(x = Year, y = mt, color = Var))+   
  geom_line(size = 1.5, alpha = 0.7)+   
  scale_color_manual(values = c("#2E86AB", "#F24236")) +  
  theme_bw() +
  facet_wrap(~Stock, ncol = 3, scales = 'free') +
  labs(x = "Year", 
       y = "mt",
       title = "TAC and ABC in the GOA") 
```

And then for the GOA.

```{r raw-specs-goa}
#| fig-width: 8
#| fig-height: 12
#for_plot_goa <- c("Pollock", "Pacific cod", "Sablefish", "Shallow-water Flatfish", "Deep-water Flatfish",  "Arrowtooth Flounder", "Flathead Sole", "Pacific Ocean Perch", "Rex Sole")

goa_specs_tot %>%   
  #filter(Stock %in% for_plot_goa) %>%
  ggplot(aes(x = Year, y = mt, color = Var))+   
  geom_line(size = 1.5, alpha = 0.7)+   
  scale_color_manual(values = c("#2E86AB", "#F24236")) +  
  theme_bw() +
  facet_wrap(~Stock, ncol = 3, scales = 'free') +
  labs(x = "Year", 
       y = "mt",
       title = "TAC and ABC in the GOA") 
```

## Patterns

### BSAI Region

Some examples:

-   Pollock in BS: Ratio fluctuates between 50% and 100% since 2010
-   Pacific Cod in BS: Nearly 100%
-   Sablefish: High (near 100%) but declining in recent years
-   Arrowtooth Flounder: Consistently low (\<25%)
-   Pacific Ocean Perch: Relatively high (\>80%)
-   Yellowfin sole: \>60%
-   Rock sole: \<40% but increasing

### GOA Region

Some examples:

-   Pollock: Consistently 100%
-   Pacific Cod: Stable at \~75% (State of Alaska allocation?)
-   Sablefish: High but declining (high recruitment)
-   Pacific Ocean Perch, Northern rockfish, shortraker, Rex sole, Deep water flatfish: 100%
-   Arrowtooth Flounder: Increasing to \>75%
-   Shallow water flatfish: 80%
-   Flathead: Increasing

## Conclusion

TAC/ABC ratios are generally higher in the GOA compared to BSAI. In the few cases where TAC \< ABC, most of the times it is flatfish. Notably, of this TAC, attainment is very low for some stocks (e.g. 3% for deep water flatfish).\
\
In ACLIM models ATTACH is used to go ABC -\> TAC -\> catch in two steps using a series of regressions. These are trained on harvest specification data from 1992-2017 (perhaps updated now), while accounting for changes in management (e.g. Amendment 80, Steller sea lion closures, etc.).
